An Improved Method for Coupling Hydrodynamics with Astrophysical Reaction Networks

### 2022-06-01 simplified_sdc_intro/paper.tex
update to latest aastex
put in the overleaf version

%\documentclass[times,modern]{aastex63}
\documentclass[times,preprint]{aastex63}
%\documentclass{aastex63}

@@

\newcommand{\Ic}{{\boldsymbol{\mathcal{I}}}}
\newcommand{\Ics}{{\mathcal{I}}}
\newcommand{\smax}{{s_\mathrm{max}}}
\newcommand{\kth}{k_\mathrm{th}}
\usepackage{bm}

@@

\newcommand{\Shydro}{{{\bf H}}}
\newcommand{\Hb}{{\bf H}}
\newcommand{\Rb}{{\bf R}}
\newcommand{\Rq}{{\bf R}}
\newcommand{\Adv}[1]{{\left [\boldsymbol{\mathcal{A}} \left(#1\right)\right]}}
\newcommand{\Advt}[1]{{\left [\boldsymbol{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advss}[1]{{\left [{\mathcal{{A}}} \left(#1\right)\right]}}
\newcommand{\Advsst}[1]{{\left [{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advs}[1]{\boldsymbol{\mathcal{A}} \left(#1\right)}

@@

\author[0000-0003-0439-4556]{M.~P.~Katz}
\affiliation{Nvidia Corp}

\author[0000-0003-1791-0265]{A.~J.~Nonaka}
\affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

\author[0000-0002-1530-781X]{A.~Harpole}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
  Stony Brook, NY 11794-3800}

\author{M.~Rasmussen}
\affiliation{Utah State University, Logan UT 84322}

@@

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute during explosive burning or at high
temperatures where nuclear statistical equilibrium is established.  We
develop a simplified approach based on the ideas of spectral deferred
corrections (SDC) coupling of explicit hydrodynamics and stiff
reaction sources as an alternative to operator splitting or the more
comprehensive SDC approach we demonstrated previously.  We apply the
new method to a double detonation problem with a moderately-sized
astrophysical nuclear reaction network and explore the timestep size
and reaction network tolerances to show that the simplified-SDC
approach compares favorably with traditional Strang operator
splitting.  This is all done in the framework of the
\castro\ hydrodynamics code, and all algorithm implementations are
freely available.
\end{abstract}

@@

Modeling astrophysical reacting flows can be challenging because of
the disparity better the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to breakdown.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  These large changes in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

The \castro\ hydrodynamics code \citep{castro,castro_joss} is used for
all of our numerical experiments.  \castro\ is a compressible
(magneto-, radiation-) hydrodynamics code built on the AMReX adaptive
mesh refinement (AMR) framework \citep{amrex_joss}.  \castro\ has been
designed to be performance portable and runs on massively parallel CPU
and GPU architectures \citep{castro_gpu}.  For hydrodynamics, the
corner transport upwind (CTU; \citealt{ppmunsplit}) method with the
piecewise parabolic method (PPM; \citealt{ppm,millercolella:2002}) is
used.  \castro\ includes self-gravity, rotation, arbitrary equations
of state and reaction networks, and has been used for modeling X-ray
bursts and different models of thermonuclear, core-collapse, and
pair-instability supernovae.  Recently in \citet{castro_sdc}, we
developed second- and fourth-order accurate methods in space and time
for coupling hydrodynamics and nuclear reaction networks based on
spectral deferred corrections (SDC) methods, and demonstrated the
method in a variety of test problems.

The time-integration approach presented here is considerably simpler
than the SDC method of \citet{castro_sdc}, but allows us to reuse the
main CTU \MarginPar{any cites to CCSE papers?}  hydrodynamics
construction and a largely similar ODE integration scheme, making this
method easier to add to existing simulation codes.  Furthermore, it
also extends to adaptive mesh refinement with subcycling in a
straightforward manner, avoiding the complications described in
\citep{mccorquodalecolella} needed to fill ghost cells when using
method-of-lines integration.  However, it is restricted to
second-order accuracy in time overall.  We term this algorithm the
``simplified-SDC method''.  In this paper we describe the overall
method and demonstrate it on a test problem and astrophysical
simulation using \castro.  All of the code to reproduce the results in
this paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include in the CTU hydrodynamics scheme.  Our conserved variables are
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}).
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$.
Defining the hydrodynamical fluxes:
\begin{equation}
  \Fb(\Uc) = \left ( \begin{array}{c}
         \rho u \\
         \rho X_k u \\
         \rho u^2 p \\
         (\rho E + p) u \\
         \rho e u \end{array}\right )
\end{equation}
We can write the system in conservative form for all state variables aside from $(\rho e)$ as:
\begin{equation}
  \ddt{\Uc} + \ddx{\Fb(\Uc)} = \Sc(\Uc)
\end{equation}
where for the special case of $(\rho e)$, we have an additional ``pdV'' term:
\begin{equation}
\ddt{(\rho e)} + \ddx{F(\rho e)} + p \ddx{u} = S(\rho e)
\end{equation}

We'll split the source term, $\Sc(\Uc)$ into pure hydrodynamic and reactive parts:
\begin{equation}
  \label{eq:cons_sources}
  \Hb(\Uc) = \left ( \begin{array}{c}
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    0 \end{array} \right )
  \qquad
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S}
  \end{array} \right )
\end{equation}
with
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc).
\end{equation}
Here, gravity, represented by the gravitational acceleration $g$
appears as a pure hydrodynamical source term.  From reactions,
$\omegadot(X_k)$ is the creation rate for species $k$
 and $\dot{S}$ is the energy generation rate per unit
mass.  We note that the internal energy ``pdV'' work is not treated as
a source term but is instead constructed with the hydrodynamical
fluxes are computed in the CTU method.
Aside from the $p\partial u/\partial x$ (``pdV'') term in the internal
energy equation, this system is in conservative form with source
terms.

We'll define the advective terms with hydrodynamic sources, $\Advs{\Uc}$ as
\begin{equation}
\Advs{\Uc} = -\ddx{\Fb(\Uc)} + \Hb(\Uc)
\end{equation}
for the general case, with the $(\rho e)$ component again having the extra
``pdV'' term:
\begin{equation}
\mathcal{A}(\rho e) = -\ddx{F(\rho e)} -p \ddx{u} + H(\rho e)
\end{equation}

To close the system, we need the equation of state of the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sc(\qb)
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{ccccc}
    u & 0 &  \rho & 0 & 0 \\
    0 & u &  0    & 0 & 0 \\
    0 & 0 &  u    & 1/\rho & 0 \\
    0 & 0 &  \Gamma_1 p & u & 0 \\
    0 & 0 &  \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $h$ is the specific enthalpy and $\Gamma_1$ is an adiabatic index,
$\Gamma_1 = d\log p/d\log\rho|_s$ at constant entropy.
The CTU+PPM algorithm uses the characteristic wave structure of $\Ab$
to collect the information that makes it to an interface over a timestep
in order to compute the fluxes through the interface.
Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sc(\qb)$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sc(\qb) = \Hb(\qb) + \Rb(\qb)
\end{equation}
with
\begin{equation}
\label{eq:prim_sources}
\Hb(\qb) = \left ( \begin{array}{c}
     0 \\
     0 \\
     g \\
     0 \\
     0 \\
   \end{array} \right )
\qquad
\Rb(\qb) = \left ( \begin{array}{c}
     0 \\
     \omegadot(X_k) \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

The CTU+PPM method for hydrodynamics is second-order accurate in space
and time.  We want to couple the reactive sources to the hydrodynamics
to be second-order in time as well.  As discussed above, nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is
traditionally employed here, and is used as a benchmark for comparison
in this paper.  We'll discuss this traditional approach next before moving
on to our new time-coupling method.

@@

In the Strang splitting~\cite{strang:1968} flavor of operator splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

\begin{figure}[t]
  \plotone{sdc_strang_flowchart_I}
  \caption{\label{fig:strang_flowchart} Flowchart of the operator split,
    Strang, time integration.  We first react for $\Delta t/2$, then
    advect for $\Delta t$, and finally react for $\Delta t/2$.  Each
    process uses the state left behind from the previous process.}
  \end{figure}

In the absence of advective terms, our reaction system appears as just
$d\Uc/dt = \Rb(\Uc)$, or:
\begin{align}
\odt{\rho} & = 0 \\
\odt{(\rho X_k)} &= \rho \omegadot(X_k) \\
\odt{(\rho u)} &= 0 \\
\odt{(\rho E)} &= \rho \Sdot
\end{align}
We can write the energy equation as:
\begin{equation}
\odt{(\rho E)} = \odt{(\rho e)} + \odt{K} = \rho \Sdot
\end{equation}
where $K$ is the kinetic energy, $K = \rho |u|^2/2$.  Since the density and velocity
are unchanged by reactions (when Strang splitting), our energy equation becomes:
\begin{equation}
\label{eq:strang:e}
\odt{(\rho e)} = \rho \odt{e} = \rho \Sdot
\end{equation}
The reaction rates are typically expressed as $\omegadot_k(\rho, T, X_k)$
when we evolve this system, which requires us to get temperature from
the equation of state each
time we need to evaluate the reactive terms.

We also typically integrate mass fraction itself, instead of partial
densities:
\begin{equation}
\label{eq:strang:X}
\odt{X_k} = \omegadot_k
\end{equation}
We integrate Equations (\ref{eq:strang:e}) and (\ref{eq:strang:X})
using an implicit ODE solver described below.

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  A variation on Strang splitting called (re-)balanced
splitting was developed in \citet{speth:2013}, but we do not explore
that here. \MarginPar{keep this sentence?}

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach (see \citealt{ppmunsplit}).  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Eq.~\ref{eq:dt:nuce_I} uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Eq.~\ref{eq:dt:nuce_II} uses the integral of the
energy generation rate over the entire step (or last Strang half).
Usually these limiters are used in a reactive manner---the next step will see
a smaller timestep based on what happened during the current step, but the
current step is still accepted, even if it violated these conditions.  Our
goal is to avoid the need for these limiters by doing the coupling of hydrodynamics
and reactions better.

\castro\ has the ability to reject a timestep if it detects a failure and retry
with smaller timesteps (subcycling to make up the original required timestep).
Among the conditions that can trigger this are density falling below zero during
advection, the ODE integration failing to converge in the implicit solve (due to
too many steps or the internal timestep falling below a minimum), or violation
of one of the timestep limiters during the step.  This means that the timestep
constraints are proactive instead of reactive.  The retry mechanism in \castro\
works with both the Strang and simplified-SDC integration scheme.  Retrying the
step is much more stringent that simply adjusting the next step, so we put a cap
on the number of retries allowed, aborting if we exceed the cap. \MarginPar{describe in more detail?}

@@

The basic idea of spectral deferred corrections \MarginPar{ref} is to express the
update as an integral and divide the time-update into a number of
discrete time nodes.  The integral is then approximated using a
quadrature rule over these time nodes and low order approximations are
used to update the state from one time node to the next.  The method
uses iteration to successively improve the solution and the ultimate
accuracy is determined by the quadrature method used to evaluate the
integral, which is done using an iteratively-lagged solution.  This is the
classic approach that we implemented in Castro in \citet{castro_sdc}.

>>>>

%\documentclass[times,modern]{aastex63}
\documentclass[times,preprint]{aastex631}
%\documentclass{aastex63}

@@

\newcommand{\Ic}{{\boldsymbol{\mathcal{I}}}}
\newcommand{\Ics}{{\mathcal{I}}}
\newcommand{\kmax}{{k_\mathrm{max}}}
\newcommand{\kth}{k_\mathrm{th}}
\usepackage{bm}

@@

\newcommand{\Gb}{{\bf G}}
\newcommand{\Rb}{{\bf R}}
\newcommand{\Rq}{{\bf R}}
\newcommand{\Adv}[1]{{\left [\boldsymbol{\mathcal{A}} \left(#1\right)\right]}}
\newcommand{\Advt}[1]{{\left [\boldsymbol{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advss}[1]{{\left [{\mathcal{{A}}} \left(#1\right)\right]}}
\newcommand{\Advsst}[1]{{\left [{\mathcal{\tilde{A}}} \left(#1\right)\right]}}
\newcommand{\Advs}[1]{\boldsymbol{\mathcal{A}} \left(#1\right)}

@@

\author[0000-0003-0439-4556]{M.~P.~Katz}
\affiliation{Dept.\ of Physics and Astronomy, Stony Brook University,
  Stony Brook, NY 11794-3800}

\author[0000-0003-1791-0265]{A.~Nonaka}
\affiliation{Center for Computational Sciences and Engineering, Lawrence Berkeley National Laboratory, Berkeley, CA  94720}

\author[0000-0002-0297-0313]{M.~Rasmussen}
\affiliation{Utah State University, Logan UT 84322}

@@

%======================================================================
% Abstract and Keywords
%======================================================================
\begin{abstract}
Reacting astrophysical flows can be challenging to model because of
the difficulty in accurately coupling hydrodynamics and reactions.
This can be particularly acute during explosive burning or at high
temperatures where nuclear statistical equilibrium is established.  We
develop a new approach based on the ideas of spectral deferred
corrections (SDC) coupling of explicit hydrodynamics and stiff
reaction sources as an alternative to operator splitting that is simpler
than the more
comprehensive SDC approach we demonstrated previously.  We apply the
new method to a double detonation problem with a moderately-sized
astrophysical nuclear reaction network and explore the timestep size
and reaction network tolerances to show that the simplified-SDC
approach provides improved coupling with decreased computational
expense compared to traditional Strang operator
splitting.  This is all done in the framework of the
\castro\ hydrodynamics code, and all algorithm implementations are
freely available.
\end{abstract}

@@

Modeling astrophysical reacting flows can be challenging because of
the disparity between the nuclear and hydrodynamics timescales.
Reaction networks tend to be stiff, requiring implicit integration
techniques to stably integrate the system~\citep{BYRNE19871}.  In contrast, compressible
hydrodynamics flows are limited by the (often much longer)
sound-crossing time over a computational cell and can be solved using
explicit time integration. Traditional methods of coupling
hydrodynamics and reactions used in astrophysics use operator
splitting---each physical process acts on the output of the previous
process in alternating fashion.  This makes it easy to use different
time-integration methods for the different physics, and to build a
simulation code in a modular way.  However, competition between the
different physical processes can cause the coupling to breakdown.
These splitting errors can lead to loss of accuracy and further time
step limitations.

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in for temperatures in excess of
$\mbox{few} \times 10^9~\mathrm{K}$.  Physically, the forward and
reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium. 
These large changes in abundances (and large alternately positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep (see, e.g., \citealt{couch:2015,rivas:2022}), but this can make simulations
prohibitively expensive.  Sometimes the burning is simply halted on a
zone-by-zone basis when NSE conditions are reached (e.g., as in
\citealt{hedet}).  The
present work focuses on networks alone, but in a follow-on paper we
will explore hybrid burning models consisting of networks and tables for
nuclear statistical equilibrium.

The \castro\ hydrodynamics code \citep{castro,castro_joss} is used for
all of our numerical experiments.  \castro\ is a compressible
(magneto-, radiation-) hydrodynamics code built on the AMReX adaptive
mesh refinement (AMR) framework \citep{amrex_joss}.  \castro\ has been
designed to be performance portable and runs on massively parallel CPU,
multicore, and GPU architectures \citep{castro_gpu}.  For hydrodynamics, the
corner transport upwind (CTU; \citealt{ppmunsplit}) method with the
piecewise parabolic method (PPM; \citealt{ppm,millercolella:2002}) is
used.  \castro\ includes self-gravity, rotation, arbitrary equations
of state and reaction networks, and has been used for modeling X-ray
bursts and different models of thermonuclear, core-collapse, and
pair-instability supernovae.  Recently in \citet{castro_sdc}, we
developed second- and fourth-order accurate in space and time
method-of-lines approaches for coupling hydrodynamics and nuclear reaction networks based on
spectral deferred corrections (SDC), and demonstrated these
methods using a variety of test problems.

The time-integration approach presented here is considerably simpler
than the SDC method of \citet{castro_sdc}, but allows us to reuse the
piecewise-linear or piecewise-parabolic CTU hydrodynamics
construction \citep{saltzman1994,millercolella:2002}
used in the the original \castro\ paper
%\MarginPar{any cites to CCSE papers? - you would want Colella90 (2D CTU), Saltzman94 (3D CTU), Miller/Colella02 (2D/3D PPM)}
and a largely similar ODE integration scheme, making this
method easier to add to existing simulation codes.  Furthermore, it
also extends to adaptive mesh refinement with subcycling in a
straightforward manner, avoiding the complications described in
\cite{mccorquodalecolella} needed to fill ghost cells when using
method-of-lines integration.  However, it is restricted to
second-order accuracy in time overall.  We term this algorithm the
``simplified-SDC method''.  In this paper we describe the overall
method and demonstrate it on a test problem and astrophysical
simulation using \castro.  All of the code to reproduce the results in
this paper are in the \castro\ GitHub
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include in the CTU hydrodynamics scheme.  Our conserved variables are
\begin{equation}
  \Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \\
           \rho e \end{array}\right )
\end{equation}
where $\rho$ is the mass density, $u$ is velocity, $E$ is specific
total energy, $p$ is the pressure, and we carry nuclear species mass
fractions, $X_k$.  The specific total
energy relates to the specific internal energy, $e$, as $E = e + u^2/2$,
and we also separately evolve $e$, as part of a dual
energy formulation (see \citealt{bryan:1995,wdmergerI}).
 The mass fractions are constrained to sum to 1, $\sum_k X_k = 1$.
Defining the hydrodynamical fluxes:
\begin{equation}
  \Fb(\Uc) = \left ( \begin{array}{c}
         \rho u \\
         \rho X_k u \\
         \rho u^2 p \\
         (\rho E + p) u \\
         \rho e u \end{array}\right )
\end{equation}
We can write the system in conservative form for all state variables aside from $(\rho e)$ as:
\begin{equation}
  \ddt{\Uc} + \ddx{\Fb(\Uc)} = \Sc(\Uc)
\end{equation}
where for the special case of $(\rho e)$, we have an additional ``$p\, dV$'' term:
\begin{equation}
\ddt{(\rho e)} + \ddx{F_{\rho e}} + p \ddx{u} = S_{\rho e}
\end{equation}
We split the source term into gravitational and reactive parts, $\Sc(\Uc) = \Gb(\Uc) + \Rb(\Uc)$, with
\begin{equation}
  \label{eq:cons_sources}
  \Gb(\Uc) = \left ( \begin{array}{c}
    0 \\
    0 \\
    \rho g \\
    \rho u g \\
    0 \end{array} \right ),
  \qquad
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \dot\omega_{X_k} \\
     0 \\
     \rho \dot{S} \\
     \rho \dot{S}
  \end{array} \right ).
\end{equation}
From reactions,
$\dot\omega_{X_k}$ is the creation rate for species $k$
 and $\dot{S}$ is the energy generation rate per unit
mass.  We note that the internal energy $p\partial u/\partial x$ (``$p\, dV$'')
work is not treated as
a source term but is instead constructed with the hydrodynamical
fluxes that are computed in the CTU method.
Aside from the ``$p\, dV$'' term in the internal
energy equation, this system is in conservative form with source
terms.

We define the advective terms with gravitational sources, $\Advs{\Uc}$, as
\begin{equation}
\Advs{\Uc} = -\ddx{\Fb(\Uc)} + \Gb(\Uc)
\end{equation}
for the general case, with the $(\rho e)$ component again having the extra
``$p\, dV$'' term:
\begin{equation}
\mathcal{A}_{\rho e} = -\ddx{F_{\rho e}} -p \ddx{u} + G_{\rho e}
\end{equation}
To close the system, we need an equation of state of the form:
\begin{equation}
p = p(\rho, {\bf X}, e).
\end{equation}
Sometimes it is preferable to work with the primitive variables:
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sc(\qb)
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{ccccc}
    u & 0 &  \rho & 0 & 0 \\
    0 & u &  0    & 0 & 0 \\
    0 & 0 &  u    & 1/\rho & 0 \\
    0 & 0 &  \Gamma_1 p & u & 0 \\
    0 & 0 &  \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $h$ is the specific enthalpy and $\Gamma_1$ is an adiabatic index,
$\Gamma_1 = d\log p/d\log\rho|_s$ at constant entropy.
The CTU+PPM algorithm uses the characteristic wave structure of $\Ab$
to collect the information that makes it to an interface over a timestep
in order to compute the fluxes through the interface.
Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sc(\qb)$, can again be decomposed into gravitational
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sc(\qb) = \Gb(\qb) + \Rb(\qb),
\end{equation}
with
\begin{equation}
\label{eq:prim_sources}
\Gb(\qb) = \left ( \begin{array}{c}
     0 \\
     0 \\
     g \\
     0 \\
     0 \\
   \end{array} \right ),
\qquad
\Rb(\qb) = \left ( \begin{array}{c}
     0 \\
     \dot\omega_{X_k} \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right ),
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Equation (25) of \cite{castro}.

The CTU+PPM method for hydrodynamics is second-order accurate in space
and time.  We want to couple the reactive sources to the hydrodynamics
to be second-order in time as well.  As discussed above, nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting (e.g., Strang) is
traditionally employed here, and is used as a benchmark for comparison
in this paper.  We discuss this traditional approach next before moving
on to our new time-coupling method.

@@

In the Strang splitting~\citep{strang:1968} flavor of operator splitting, we first integrate the system with reactions terms only (no advection)
over $\Delta t/2$, then integrate the advection terms only (no reactions) over $\Delta t$,
and finally integrate the reaction terms only over $\Delta t/2$.  This flow is illustrated
in Figure~\ref{fig:strang_flowchart}.

\begin{figure}[t]
  \plotone{sdc_strang_flowchart_I}
  \caption{\label{fig:strang_flowchart} Flowchart of the operator split (Strang) time integration.  We first react for $\Delta t/2$, then
    advect for $\Delta t$, and finally react for $\Delta t/2$.  Each
    process uses the state left behind from the previous process.}
  \end{figure}

In the absence of advective terms, our reaction system appears as just
$d\Uc/dt = \Rb(\Uc)$, or:
\begin{align}
\odt{\rho} & = 0 \\
\odt{(\rho X_k)} &= \rho \dot\omega_{X_k} \\
\odt{(\rho u)} &= 0 \\
\odt{(\rho E)} &= \rho \Sdot
\end{align}
Notice that in the Strang formulation density is held constant when reacting.
We can write the energy equation as:
\begin{equation}
\odt{(\rho E)} = \odt{(\rho e)} + \odt{K} = \rho \Sdot
\end{equation}
where $K$ is the kinetic energy, $K = \rho |u|^2/2$.  Since the density and velocity
are unchanged by reactions (when Strang splitting), our energy equation becomes:
\begin{equation}
\label{eq:strang:e}
\odt{(\rho e)} = \rho \odt{e} = \rho \Sdot
\end{equation}
The reaction rates are typically expressed as $\omegadot_k(\rho, T, {\bf X})$
when we evolve this system, which requires us to get temperature from
the equation of state each
time we need to evaluate the reactive terms.

We also typically integrate mass fractions, instead of partial
densities:
\begin{equation}
\label{eq:strang:X}
\odt{X_k} = \omegadot_k
\end{equation}
We integrate Equations (\ref{eq:strang:e}) and (\ref{eq:strang:X})
using an implicit ODE solver described below.

We explored this and other approaches (including not evolving an
energy / temperature equation during the reaction step) in
\citet{strang_rnaas} and showed that the above formulation gets second
order convergence and that in particular, integrating an energy / temperature
equation is needed to get second order accuracy (see also
\citealt{muller:1986} for a discussion on stability).  We note however
that some astrophysical simulation codes only evolve the species
equations.  For very strong reactions, when using Strang splitting the
state can drift significantly off of the smooth solution to the
coupled reactive hydrodynamics equations, as shown graphically in
\citet{astronum:2018} (using an earlier version of the present
algorithm).  
%A variation on Strang splitting called (re-)balanced
%splitting was developed in \citet{speth:2013}, but we do not explore
%that here. \MarginPar{keep this sentence? MK: Don't think it's needed.}

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling, the advective CFL condition is unity) as
compared to a method-of-lines approach; see \cite{ppmunsplit}.  In
addition to the standard CFL timestep limiter for explicit
hydrodynamics, timestep limiters based on
the energy generation or abundance changes over a timestep, such as
those introduced in \cite{prometheus}, are often used.  An energy limiter takes the
form of either:
\begin{equation}
\label{eq:dt:nuce_I}
\Delta t \le f_e\, \min_{i} \left\{\frac{e^n_{i}}{|\dot{S}_{i}^n|}\right\} \\
\end{equation}
or
\begin{equation}
\label{eq:dt:nuce_II}
\Delta t \le f_e\, \min_{i} \left\{ \frac{e^n_i}{|\int \dot{S}_i(t) dt|} \right\}
\end{equation}
where $i$ is the zone index and $f_e$ is a parameter used to control
the allowed change (e.g., $f_e = 0.5$ will limit the step such that
reactions can only increase the internal energy by 50\%).  The two
formulations differ in that Equation (\ref{eq:dt:nuce_I}) uses the
instantaneous evaluation of the energy generation rate at the start of
the timestep while Equation (\ref{eq:dt:nuce_II}) uses the integral of the
energy generation rate over the entire step (or last Strang half).
Usually these limiters are used in a reactive manner---the next step will see
a smaller timestep based on what happened during the current step, but the
current step is still accepted, even if it violated these conditions.  Our
goal is to avoid the need for these limiters by improving the coupling of hydrodynamics
and reactions.

\castro\ has the ability to reject a timestep if it detects a failure and retry
with smaller timesteps (subcycling to make up the original required timestep).
Among the conditions that can trigger this are density falling below zero during
advection, the ODE integration failing to converge in the implicit solve (due to
too many steps or the internal timestep falling below a minimum), or violation
of one of the timestep limiters during the step.  This means that the timestep
constraints are proactive instead of reactive.  The retry mechanism in \castro\
works with both the Strang and simplified-SDC integration scheme.  Retrying the
step is much more stringent that simply adjusting the next step, so we put a cap
on the number of retries allowed, aborting if we exceed the cap. 
%\MarginPar{describe in more detail? MK: While I do want us to evangelize the retry scheme, it %seems slightly out of place in this paper.  MZ: I do note later that the retry kicked in for %both Strang and SDC, so that's why I added it here.}

@@

The basic idea of spectral deferred corrections \citep{dutt:2000} is to divide each time step into sub-steps defined by high-order quadrature points (e.g.~Gauss-Lobatto) and iteratively correct the solution at each temporal node in order to reduce the integration error.
Each correction equation can be formulated using low-order integration techniques, such as forward or backward-Euler, with additional terms on each right-hand-side arising from integrating the residual over the sub-step from the previous iteration using high-order temporal integration.
Each iteration over all sub-steps increases the overall order of accuracy of the method by one, up to the underlying order of accuracy of the quadrature over the entire time step.
A number of SDC approaches have been developed
over the years targeting tighter multiphysics coupling, such as 
\cite{bourlioux2003high,SDC-old}.  In these approaches, different physical processes can be treated with different approaches (such as forward-Euler for advection and backward-Euler for reactions), and furthermore different physical processes can be sub-stepped between temporal nodes.
